---
title: "DuckDb vs SQLite"
output: html_document
date: "2023-06-02"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# DuckDb versus SQLite

### Structure and Performance

|OLTP (Online Transaction Processing) - SQLite|OLAP (Online Analytical Processing) - DuckDb|
|---------------|---------------|
|row oriented storage (high processing overhead)|column oriented storage (low processing overhead)|
|Optimized for a high volume of simple transactions like data insertion, deletion, and updates)|Optimized for complex analytical queries that involve calculations, summarizations, and comparisons across multiple dimensions|

</br>
Lets do a basic performance test for read and write transactions.
``` {r, eval = FALSE}

conn_s <- DBI::dbConnect(RSQLite::SQLite(), dbname = ":memory:")
conn_d <- DBI::dbConnect(duckdb::duckdb(), dbdir = ":memory:")

data <- data.frame(
  id = 1:10000000,
  value = rnorm(10000000)
)
sqlite_write <- system.time(DBI::dbWriteTable(conn_s, "data", data))
sqlite_read <- system.time(DBI::dbGetQuery(conn_s, "SELECT * FROM data"))

duckdb_write <- system.time(DBI::dbWriteTable(conn_d, "data", data))
duckdb_read <- system.time(DBI::dbGetQuery(conn_d, "SELECT * FROM data"))

paste("SQLite write time:", round(sqlite_write[3], 3), "seconds")
paste("SQLite read time:", round(sqlite_read[3], 3), "seconds")

paste("DuckDB write time:", round(duckdb_write[3], 3), "seconds")
paste("DuckDB read time:", round(duckdb_read[3], 3), "seconds")


```

```{r, eval = FALSE}
## [1] "SQLite write time: 2.529 seconds"
## [1] "SQLite read time: 2.318 seconds"

## [1] "DuckDB write time: 0.192 seconds"
## [1] "DuckDB read time: 0.029 seconds"
```

### Supported data types

DuckDb supports all of the standard data types in SQL (including extended types) and more:
`BOOLEAN`, `TINYINT`, `SMALLINT`, `INTEGER`, `BIGINT`, `FLOAT`, `DOUBLE`, `VARCHAR`, `DATE`, `TIME`, `TIMESTAMP`, `INTERVAL`, `HUGEINT`.

Notably, DuckDb aslo supports some nested data types: `LIST`, `STRUCT`, `MAP` and `UNION`.

Here we will demonstrate the behavior for the subset of types we use most commonly use.


```{r, results = 'hide'}
data <- data.frame(
  logical = TRUE,
  integer = 1L,
  real = 1.1,
  character = "a",
  blob = blob::as_blob(NA),
  date = as.Date("2023-01-01"),
  time = dttr2::dtt_time("1:00:00"),
  datetime = dttr2::dtt_date_time("2023-01-01 01:00:00")
)

table_definition <- "CREATE TABLE data (
       logical BOOLEAN,
       integer INTEGER,
       real REAL,
       character TEXT,
       blob BLOB,
       date DATE,
       time TIME,
       datetime TIMESTAMP
)"

# SQL
conn_s <- DBI::dbConnect(
  RSQLite::SQLite(), dbname = ":memory:", extended_types = TRUE
  )

DBI::dbExecute(conn_s, table_definition)
DBI::dbAppendTable(conn_s, "data", data)

# DuckDb
conn_d <- DBI::dbConnect(duckdb::duckdb(), dbdir = ":memory:")

DBI::dbExecute(conn_d, table_definition)
DBI::dbAppendTable(conn_d, "data", data)

```
```{r}
# compare
sql <- DBI::dbReadTable(conn_s, "data")
duck <- DBI::dbReadTable(conn_d, "data")

DBI::dbDisconnect(conn_s, shutdown = TRUE)
DBI::dbDisconnect(conn_d, shutdown = TRUE)

str(sql)
str(duck)

```

</br>
Notably, there are differences in the class of objects returned for times, blobs, and logical data types

|Defined Type|Class Returned SQL|Class Returned DuckDb|Explanation|
|------------|------------|------------|--------------------------------------|
|__boolean__|__integer__|__logical__|SQL returns a `1` or `0` equivalent of `TRUE` or `FALSE`|
|integer|integer|integer||
|real|numeric|numeric||
|text|character|character||
|__blob__|__blob__, __vctrs_list_of__, __vctrs_vctr__, __list__|__list__|SQL returns a proper `blob` object while DuckDB returns a `raw` vector|
|date|Date|Date|
|__time__|__hms__, __difftime__|__difftime__|There is curated support for `hms` object with RSQLite, while DuckDB just returns the time difference from the beginning of the day|
|timestamp|POSIXct, POSIXt|POSIXct, POSIXt|note: both engines convert datetimes to utc then store as numeric in the database. The returned datetime is in UTC|

</br>

### Type Enforcement: DuckDb

DuckDB natively enforces the types defined in tables by erroring when the data provided is incompatible with the column definition.

```{r}
# Define invalid values for each column type
invalid_values <- 
  list(
  "BOOLEAN" = "a",
  "INTEGER" = "a",
  "REAL" = "a",
  "BLOB" = 1L,
  "DATE" = "a",
  "DATE" = 1,
  "DATE" = "2023-13-01",
  "TIME" = "a",
  "TIME" = 1,
  "TIME" = "25:00:00",
  "TIMESTAMP" = "a",
  "TIMESTAMP" = 1,
  "TIMESTAMP" = "2023-01-01 25:00:00"
  )

column_types <- names(invalid_values)

msgs <- c()
i <- 1
for (i in seq_along(column_types)) {

  conn <- DBI::dbConnect(duckdb::duckdb(), dbdir = ":memory:")
  column_type <- column_types[i]

  # generate table with each column type
  sql <- paste0("CREATE TABLE data (",
                "col ", column_type,
                ")")
  DBI::dbExecute(conn, sql)

  # try to insert invalid data
  err <- try(
    expr = DBI::dbAppendTable(conn, "data", data.frame(col = invalid_values[[i]])),
    silent = TRUE
  )
  DBI::dbDisconnect(conn, shutdown = TRUE)

  msgs <- c(msgs, err[1])
}

# clean up error msgs
msgs <- gsub("Error : rapi_execute: Failed to run query\nError: Conversion Error: ", "", msgs)

error_summary <-
  tibble::tibble(
  db_defined_type = column_types,
  supplied_r_class = as.character(sapply(invalid_values, class)),
  supplied_r_value = as.character(invalid_values),
  error = msgs
  )

knitr::kable(error_summary, format = "html")
```



Note: No errors occur for TEXT columns as any R type can and will be coerced (even blobs)

### Type Enforcement: SQL

SQL does not enforce data types.

``` {r}

table_definition <- 
"CREATE TABLE data (
       logical BOOLEAN,
       integer INTEGER,
       real REAL,
       blob BLOB,
       date DATE,
       time TIME,
       datetime TIMESTAMP
)"

invalid_data <-
  data.frame(
  logical = "a", 
  integer = "a",
  real = "a",
  blob = 1L,
  date = "a",
  time = "a",
  datetime = "a"
  )

conn <- DBI::dbConnect(
  RSQLite::SQLite(), dbname = ":memory:", extended_types = TRUE
)

DBI::dbExecute(conn, table_definition)
DBI::dbAppendTable(conn, "data", invalid_data)

```

All data is written to the database with no errors and without changing the table definition.

### Geometries

Due to the differences in the way BLOBs are stored between the two platforms, slightly different operations are required to get back the original object after reading from the database.


```{r, results = 'hide'}
crs <- 4326

data <- data.frame(
  geometry = sf::st_sfc(sf::st_point(c(1, 1)), crs = crs)
)

# convert geometry to blob
data$geometry <- sf::st_as_binary(data$geometry)


table_definition <- "CREATE TABLE data (geometry BLOB)"

# SQL
conn_s <- DBI::dbConnect(
  RSQLite::SQLite(), dbname = ":memory:", extended_types = TRUE
)

DBI::dbExecute(conn_s, table_definition)
DBI::dbAppendTable(conn_s, "data", data)

# DuckDb
conn_d <- DBI::dbConnect(duckdb::duckdb(), dbdir = ":memory:")

DBI::dbExecute(conn_d, table_definition)
DBI::dbAppendTable(conn_d, "data", data)
```
```{r}
# compare
sql <- DBI::dbReadTable(conn_s, "data")
duck <- DBI::dbReadTable(conn_d, "data")

# simple conversion to sf column and activate
sql$geometry <- sf::st_as_sfc(sql$geometry, crs = crs)
sql <- sf::st_as_sf(sql)

# additional step needed to convert the raw vector to a blob before st_as_sfc
# will correctly convert it to an sf object
duck$geometry <- blob::as_blob(duck$geometry)
duck$geometry <- sf::st_as_sfc(duck$geometry, crs = crs)
duck <- sf::st_as_sf(duck)

DBI::dbDisconnect(conn_s, shutdown = TRUE)
DBI::dbDisconnect(conn_d, shutdown = TRUE)

identical(sql, duck)

```
