---
title: "DuckDb vs SQLite"
output: html_document
date: "2023-06-02"
---

# DuckDb versus SQLite

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Supported data types

DuckDb supports all of the standard data types in SQL as well as a couple more:
`BOOLEAN`, `TINYINT`, `SMALLINT`, `INTEGER`, `BIGINT`, `FLOAT`, `DOUBLE`, `VARCHAR`, `DATE`, `TIME`, `TIMESTAMP`, `INTERVAL`, `HUGEINT`.
Here we will demonstrate the behavior for the subset of types we use most commonly use.

The implementation is effectively identical for our purposes as there is an R engine equivalent of RSQLite for DuckDb called `duckdb`.


```{r, results = 'hide'}
data <- data.frame(
  logical = TRUE,
  integer = 1L,
  real = 1.1,
  character = "a",
  blob = blob::as_blob(NA),
  date = as.Date("2023-01-01"),
  time = dttr2::dtt_time("1:00:00"),
  datetime = dttr2::dtt_date_time("2023-01-01 01:00:00")
)

table_definition <- "CREATE TABLE data (
       logical BOOLEAN,
       integer INTEGER,
       real REAL,
       character TEXT,
       blob BLOB,
       date DATE,
       time TIME,
       datetime TIMESTAMP
)"

# SQL
conn_s <- DBI::dbConnect(
  RSQLite::SQLite(), dbname = ":memory:", extended_types = TRUE
  )

DBI::dbExecute(conn_s, table_definition)
DBI::dbAppendTable(conn_s, "data", data)

# DuckDb
conn_d <- DBI::dbConnect(duckdb::duckdb(), dbname = ":memory:")

DBI::dbExecute(conn_d, table_definition)
DBI::dbAppendTable(conn_d, "data", data)

```
```{r}
# compare
sql <- DBI::dbReadTable(conn_s, "data")
duck <- DBI::dbReadTable(conn_d, "data")

DBI::dbDisconnect(conn_s, shutdown = TRUE)
DBI::dbDisconnect(conn_d, shutdown = TRUE)

str(sql)
str(duck)

```

</br>
Notably, there are differences in the class of objects returned for times, blobs, and logical data types

|Defined Type|Returned SQL|Returned DuckDb|Explanation|
|------------|------------|------------|--------------------------------------|
|__boolean__|__integer__|__logical__|SQL returns a `1` or `0` equivalent of `TRUE` or `FALSE`|
|integer|integer|integer||
|real|numeric|numeric||
|text|character|character||
|__blob__|__blob__, __vctrs_list_of__, __vctrs_vctr__, __list__|__list__|SQL returns a proper `blob` object while DuckDB returns a `raw` vector|
|date|Date|Date|
|__time__|__hms__, __difftime__|__difftime__|There is curated support for `hms` object with RSQLite, while DuckDB just returns the time difference from the beginning of the day|
|timestamp|POSIXct, POSIXt|POSIXct, POSIXt|note: both engines convert datetimes to utc then store as numeric in the database. The returned datetime is in UTC|

</br>

### Type Enforcement: DuckDb

DuckDB natively enforces the types defined in tables by erroring when the data provided is incompatible with the column definition.

```{r}
# Define invalid values for each column type
invalid_values <- list(
  "BOOLEAN" = "a",
  "INTEGER" = "a",
  "REAL" = "a",
  "BLOB" = 1L,
  "DATE" = "a",
  "DATE" = 1,
  "DATE" = "2023-13-01",
  "TIME" = "a",
  "TIME" = 1,
  "TIME" = "25:00:00",
  "TIMESTAMP" = "a",
  "TIMESTAMP" = 1,
  "TIMESTAMP" = "2023-01-01 25:00:00"
)

column_types <- names(invalid_values)

msgs <- c()
i <- 1
for (i in seq_along(column_types)) {

  conn <- DBI::dbConnect(duckdb::duckdb(), dbname = ":memory:")
  column_type <- column_types[i]

  # generate table with each column type
  sql <- paste0("CREATE TABLE data (",
                "col ", column_type,
                ")")
  DBI::dbExecute(conn, sql)

  # try to insert invalid data
  err <- try(
    expr = DBI::dbAppendTable(conn, "data", data.frame(col = invalid_values[[i]])),
    silent = TRUE
  )
  DBI::dbDisconnect(conn, shutdown = TRUE)

  msgs <- c(msgs, err[1])
}

# clean up error msgs
msgs <- gsub("Error : rapi_execute: Failed to run query\nError: Conversion Error: ", "", msgs)

error_summary <- tibble::tibble(
  db_column_type = column_types,
  supplied_r_class = as.character(sapply(invalid_values, class)),
  supplied_r_value = as.character(invalid_values),
  error = msgs
)

print(error_summary)

```
Note: No errors occur for TEXT columns as any R type can and will be coerced (even blobs)

### Type Enforcement: SQL

SQL does not enforce data types.

``` {r}

invalid_values <- list(
  "BOOLEAN" = "a",
  "INTEGER" = "a",
  "REAL" = "a",
  "BLOB" = 1L,
  "DATE" = "a",
  "DATE" = 1,
  "DATE" = "2023-13-01",
  "TIME" = "a",
  "TIME" = 1,
  "TIME" = "25:00:00",
  "TIMESTAMP" = "a",
  "TIMESTAMP" = 1,
  "TIMESTAMP" = "2023-01-01 25:00:00"
)

column_types <- names(invalid_values)

results <- list()
i <- 1
for (i in seq_along(column_types)) {
  
  conn <- DBI::dbConnect(RSQLite::SQLite(), dbname = ":memory:")
  column_type <- column_types[i]
  
  # generate table with each column type
  sql <- paste0("CREATE TABLE data (",
                "col ", column_type,
                ")")
  DBI::dbExecute(conn, sql)
  
  # try to insert invalid data
  DBI::dbAppendTable(conn, "data", data.frame(col = invalid_values[[i]]))
  data <- DBI::dbReadTable(conn, "data")
  print(data$col)
  
}

```


### Other Notes
DuckDB is not compatible with DB browser for active viewing of the database tables, still looking for the best alternative for viewing.

