--- 
title: "duck db docs"
author: "Evan Amies-Galonski"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: rstudio/bookdown-demo
description: "Documentation to outline the features and limitations of duck db and its advantages over SQLite."
---
# Introduction

DuckDb is a relational database management system that supports the Structured Query Language (SQL).

### Installation

DuckDb can be installed with brew or via this link:
https://github.com/duckdb/duckdb/releases/download/v0.8.0/duckdb_cli-osx-universal.zip

```{r eval=FALSE}
$ brew install duckdb
```

DuckDb has an R engine equivalent of RSQLite for DuckDb called `duckdb`.
The implementation is effectively identical for our purposes as DuckDb interfaces with DBI just like RSQLite.

```{r eval=FALSE}
install.packages("duckdb")

# RSQLite
conn <- DBI::dbConnect(RSQLite::SQLite(), dbname = ":memory:", extended_types = TRUE)

# DuckDb
conn <- DBI::dbConnect(duckdb::duckdb(), dbname = ":memory:")

```
